~\Large Stickle!

### 2023-05-23 manuscript/paper.Rmd
Update post-imputation model specifications.

# Models  {#sec:models}
## Imputation Model 
Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish, $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times K_{cont}$ matrix of the $K_{cont}$ continuous phenotypes of interest, and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K_{disc}$ matrix of the $K_{disc}$ discrete phenotypes of interest.  Because the gender of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant gender and missing fossil gender, respectively.

@@

## Completed Data Model 
For each of the $M = 100$ imputed datasets, we fit an OU model for different phenotypes of interest.  For the $i$-th observed fossil in year $t$ ($t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$), let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K_{cont} \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $K_{disc} \times 1$ vector of discrete phenotypes for.  In addition,

\begin{align}
X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:x}
\end{align}

Look at \ref{eq:x}

For $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.  For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for is defined as

\begin{align}
u_{g,t} \iid \mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\label{eq:ar1}
\end{align}

$\kappa_g$ represents the gender-specific correlation between the previous and current time periods' random effects.  In addition, $\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h$ for $h \in \mathbb{N}$.

We also set

\begin{align}
u_{g,1} \iid \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\label{eq:u1}
\end{align}

This model choice is to make the distribution of the random effects stationary, i.e. $p(\mu_{g,s}) = p(\mu_{g,t})$ for $s,t \in \{1,\dots,T\}$.

Priors:

\begin{align}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\theta_g & \iid \mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\label{eq:priors}
\end{align}

>>>>

# Models  {#sec:models}
## Imputation Model 
Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish, $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ continuous phenotypes of interest, and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times L$ matrix of the $L$ discrete phenotypes of interest.  Because the gender of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant gender and missing fossil gender, respectively.

@@

## Completed Data Model 
For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).

Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume
\begin{align}
{X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right. \\
{X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:X}
\end{align}
where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.  We further set \begin{align}
\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.  We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUprocess) where we assume $u_{k,gT}$ have a marginal mean of 0. Because each time period is ~1000 years, we will discretize the OU process without loss of generality.  For $t = 2,\ldots,T$, we assume
\begin{align}
u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
\label{eq:u_ar1}
\end{align}

$\kappa_{k,g}$ represents the correlation between $\mu_{k,gt}$ and $\mu_{k,g(t+1)}$, and it also follows that $\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h$ for $h \in \mathbb{N}$.  In addition, we assume
\begin{align}
u_{k,g1} \iid \mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right).
\label{eq:u1}
\end{align}
This model choice is to preserve stationarity; i.e. $p(u_{k,gs}) = p(u_{k,gt})$ for $s \neq t$.

For the discrete phenotypes, we will assume
\begin{align}
{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
\label{eq:Y}
\end{align}
for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set
\begin{align}
\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
\label{eq:lambda}
\end{align}
We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume
\begin{align}
v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
\label{eq:v_ar1}
\end{align}
and
\begin{align}
v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
\label{eq:v1}
\end{align}

Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, we analyze the data via a Bayesian analysis.  Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}).

Priors: For $k = 1,\cdots,K$ and $l = 1,\cdots,L$,

\begin{align}
\sigma_k & \iid \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
\tau_k & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\kappa_k & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\beta_k & \iid \mathcal{N}(0,5) \\
\theta_{k,g} & \iid \mathcal{N}(0,10000)
\omega_l & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\phi_l & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\gamma_{l,g} & \iid \mathcal{N}(0,10000)
\label{eq:priors}
\end{align}

### 2023-05-23 manuscript/paper.tex
Update post-imputation model specifications.

\hypertarget{imputation-model}{%
\subsection{Imputation model}\label{imputation-model}}

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish and
\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times k\) matrix
of the \(k\) phenotypes of interest of the fish. Because
\(\boldsymbol{W}\) contains missing data, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T\)
where \(\boldsymbol{W}_{obs}\) and \(\boldsymbol{W}_{mis}\) are the
observed and missing data of the covariate gender, respectively. In this
setting, \(\boldsymbol{W}_{obs}\) is a \(n_{extant} \times 1\) vector
and \(\boldsymbol{W}_{obs}\) is a \(n_{fossil} \times 1\) vector,
perfectly corresponding to the observable gender of the modern extants
and the unobservable gender of the fossil data, respectively.

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{mis}|\boldsymbol{W}_{obs}, \boldsymbol{X})\) using
the multiple imputation with chained equations (MICE) algorithm (Buuren
and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) with
predictive mean matching. The imputation algorithm is run to obtain a
total of \(M = 100\) completed datasets.

\hypertarget{ou-model}{%
\subsection{OU model}\label{ou-model}}

After imputing sex, we fit an OU model for different phenotypes of
interest. WLOG we choose one of the columns of \(X\) to be the target
phenotype and our analysis is repeated for all phenotypes of interest.

Let \(W_{ti}\) and \(X_{ti}\) be the gender and phenotype of interest
for the \(i\)-th observation in year \(t\) for \(t = 1, \ldots, 18\) and
\(i = 1\ldots,n_{t}\). We assume that

\begin{align}
X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:x}
\end{align}

Look at \ref{eq:x}

For \(g \in \{f,m\}\), \(\theta_g\), is the defined as the overall mean
of the phenotype of interest for each gender, and \(u_{gt}\) are the
gender specific random effects, measuring the difference between the
gender and time-specific mean and the overall gender-specific mean,
which we fit to an autoregressive process of lag 1, which is the
discretization of an OU process. For \(t = 2,\ldots,T\) and
\(g \in \{f,m\}\), the AR1 model for is defined as

\begin{align}
u_{g,t} \overset{iid}{\sim}\mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\label{eq:ar1}
\end{align}

\(\kappa_g\) represents the gender-specific correlation between the
previous and current time periods' random effects. In addition,
\(\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h\) for
\(h \in \mathbb{N}\).

We also set

\begin{align}
u_{g,1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\label{eq:u1}
\end{align}

This model choice is to make the distribution of the random effects
stationary, i.e.~\(p(\mu_{g,s}) = p(\mu_{g,t})\) for
\(s,t \in \{1,\dots,T\}\).

Priors:

\begin{align}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\theta_g & \overset{iid}{\sim}\mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\label{eq:priors}
\end{align}

>>>>

\hypertarget{imputation-model}{%
\subsection{Imputation Model}\label{imputation-model}}

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish,
\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
of the \(K\) continuous phenotypes of interest, and \(\boldsymbol{Y}\)
be an \((n_{extant} + n_{fossil}) \times L\) matrix of the \(L\)
discrete phenotypes of interest. Because the gender of the fossilized
stickleback fish is unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant gender and missing fossil gender, respectively.

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})\)
using the multiple imputation with chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. The imputation algorithm is run to obtain
a total of \(M = 100\) completed datasets.

\hypertarget{completed-data-model}{%
\subsection{Completed Data Model}\label{completed-data-model}}

For a given imputed dataset, let \(W_{ti}\) be the imputed gender,
\(\boldsymbol{X}_{ti}\) be the \(K \times 1\) vector of continuous
phenotypes, and \(\boldsymbol{Y}_{ti}\) be the \(L \times 1\) vector of
discrete phenotypes for \(t = 1, \ldots, T\) and \(i = 1\ldots,n_{t}\)).

Define the last element of \(\boldsymbol{X}_{ti}\) (\(X_{K,ti}\)) as the
length of the stickleback. For \(k = 1,\cdots,K - 1\), we will assume
\begin{align}
{X}_{k,ti} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right. \\
{X}_{K,ti} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:X}
\end{align} where \(\mu_{k,ft}\) and \(\mu_{k,mt}\) represent the
time-\(t\) specific mean of phenotype \(k\) for female and male
stickleback fish, respectively, and \(\beta_k\) is an additional
parameter to account for the biological phenomenon that other continuous
phenotypes for a particular animal are positively correlated with their
length \textbf{CITATION NEEDED}. We further set \begin{align}
\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
\label{eq:mu}
\end{align} for \(g \in \{f,m\}\) where \(\theta_{k,g}\) is the overall
mean of phenotype \(X_k\) for each gender, and \(u_{k,gt}\) measures the
difference between \(\mu_{k,gt}\) and \(\theta_{k,g}\). We then fit
\(\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}\) to an
Ornstein-Uhlenback (OU) process
((\protect\hyperlink{ref-OUprocess}{\textbf{OUprocess?}})) where we
assume \(u_{k,gT}\) have a marginal mean of 0. Because each time period
is \textasciitilde1000 years, we will discretize the OU process without
loss of generality. For \(t = 2,\ldots,T\), we assume \begin{align}
u_{k,gt} \overset{iid}{\sim}\mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
\label{eq:u_ar1}
\end{align}

\(\kappa_{k,g}\) represents the correlation between \(\mu_{k,gt}\) and
\(\mu_{k,g(t+1)}\), and it also follows that
\(\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h\) for \(h \in \mathbb{N}\).
In addition, we assume \begin{align}
u_{k,g1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right).
\label{eq:u1}
\end{align} This model choice is to preserve stationarity;
i.e.~\(p(u_{k,gs}) = p(u_{k,gt})\) for \(s \neq t\).

For the discrete phenotypes, we will assume \begin{align}
{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
\label{eq:Y}
\end{align} for \(l = 1,\cdots,L\). Similar to the continuous
phenotypes, we set \begin{align}
\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
\label{eq:lambda}
\end{align} We use the log function to allow \(v_{l,gt}\) to take all
values on the real number line so we can properly fit an OU process to
these effects. We further assume \begin{align}
v_{l,gt} \overset{iid}{\sim}\mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
\label{eq:v_ar1}
\end{align} and \begin{align}
v_{l,g1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
\label{eq:v1}
\end{align}

Because we are fitting a dataset with a stochastic structure on the
means of the phenotypes, we analyze the data via a Bayesian analysis.
Bayesian data analysis is also more naturally used when we have to
impute data (\textbf{CITATION}).

Priors: For \(k = 1,\cdots,K\) and \(l = 1,\cdots,L\),

\begin{align}
\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
\tau_k & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\kappa_k & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\beta_k & \overset{iid}{\sim}\mathcal{N}(0,5) \\
\theta_{k,g} & \overset{iid}{\sim}\mathcal{N}(0,10000)
\omega_l & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\phi_l & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\gamma_{l,g} & \overset{iid}{\sim}\mathcal{N}(0,10000)
\label{eq:priors}
\end{align}
